---
title: "Homework 4 BME590"
author: "Altaful Amin"
date: "10/3/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Question 1: Get Diabetes Data Set
```{r}
#1.b. reading data into R

data <- read.csv("diabetes_data.csv", header =TRUE)
```


```{r}
#1.c. Structure of data frame

str(data)
```



```{r}
#1.c. number of variables(columns)

ncol(data)
```
# 50 variables in the dataframe

```{r}
#1.c. type of data in readmitted column

typeof(data$readmitted)


```
#integer data in readmitted column of data frame




Question 2: Preparing Data Set
```{r}
#2.b. Changing levels to "0" and "1" in readmitted
#copy of dataframe
tdata <- data

library(plyr)

#change levels in copied dataframe
levels(tdata$readmitted)[3] <- 0
levels(tdata$readmitted)[2] <- 1
levels(tdata$readmitted)[1] <- 1
```


```{r}

#2.c. 
#new data set with only the variables of interest
newdata <- subset(tdata, select = c(readmitted, num_lab_procedures, number_diagnoses, num_medications))

#Changing data type of readmitted column to numeric
newdata$readmitted <- as.numeric(levels(newdata$readmitted))[as.integer(newdata$readmitted)]


```

```{r}
library(ggplot2)
library(cowplot)
library(dplyr)
library(grid)
```



#Question 3: Principal COmponent Analysis
```{r}
#3.a. scatterplot matrix of variables of interest
library(GGally)

p<- ggpairs(newdata)
print(p)
```
# A positive correlation exists between all the variables.
The readmitted data has spikes at 0 and 1 as expected.
The no. of lab procedures has a distribution centered around 45.
The no. of diagnoses has multiple spikes with the highest being at 8.
The no.of medications is skewed left.



```{r}
#3.b. principal components of the data
newdata.pca <- princomp(newdata, cor = TRUE)

```




```{r}
#3.c. summary of principal components
summary(newdata.pca)
```



```{r}
#3.d. loadings to show all values
print(loadings(newdata.pca))
```




```{r}
#3.e. scree plot of pca data
screeplot(newdata.pca)
```

```{r}
#3.f. biplot of pca data
biplot(newdata.pca)
```

#3.g. Component 1 explains 37.126% of the total variation of the data.
#     Component 2 explains 24.873% of the total variation of the data.
#     Component 3 explains 20.734% of the total variation of the data.
     
     
#3.h. We need 3 components to explain at least 75% of the total variability 
#    of the data. Using the first 3 components, we can explain
#    82.733% of the  total variation of the data. 





#4 Visualizing Principal Components
```{r}
#4.a. 2d plots of PC1 against PC2, PC1 against PC3, and PC2 against PC3


#isolating the principal components data
PCtotal <- as.data.frame(newdata.pca$scores)


#adding readmitted column to PCA data frame
PCtotal[,5] <- newdata$readmitted

names(PCtotal) <- c("PC1","PC2","PC3","PC4","Readmitted")




```



```{r}
#generating the 2d plots

p1 <- ggplot(PCtotal, aes(x= PC1, y=PC2, color = Readmitted)) + geom_point()
p2 <- ggplot(PCtotal, aes(x= PC1, y=PC3, color = Readmitted)) + geom_point()
p3 <- ggplot(PCtotal, aes(x= PC2, y=PC3, color = Readmitted)) + geom_point()

plot_grid(p1,p2,p3, labels = "AUTO")

```




```{r}
 #4.b. 3d plot of principal components that you can rotate

library(car)
library(rgl)


plot3d(PCtotal$PC1, PCtotal$PC2, PCtotal$PC3, type="s", size=1, lit=TRUE,main = "PC1 VS PC2 VS PC3",sub="3-D Plot")


#plot opens on new window and can be rotated

```

#4.c. On the PC1 vs PC2 plot, PC1 has more variability than PC2. 
     On the PC1 vs PC3 plot, PC1 has more variability than PC3.
     On the PC2 vs PC3 plot, PC3 has more variability than PC1.
 From the loadings data, we make the following observations: 
 PC1 has a lower coefficient of readmitted compared to PC2 and PC3.
  PC2 has a lower positive coefficient of number of lab procedures compared 
  to PC1 and PC3 which have a higher negative coefficient.
  PC2 also has a smaller coefficient for number of diagnosis compared to
  PC1 and PC3. 
  PC3 is not affected by number of medications. 
  The scree plot shows us that PC1 accounts for the most variation. 



                  
